---
title: "Evaluating Short-Term Restoration Approaches for Degraded Grasslands in Kenya"
author: Ridhin Paul
format: 
  html:
    embed-resources: true
editor: visual
---

## Summary

This project evaluates the short-term effectiveness of restoration methods for degraded grasslands in Kenya. The treatments include **Tilage (T)**, the addition of native **soil inoculum (S)** and **Manure (M)**. "*A grazing exclosure was installed at the campus of the Kabianga University in Kenya in April 2022, when restoration treatments were laid out and baseline soil samples taken and analysed for key soil physical, chemical and biological properties. In November 2022, fresh soil samples were taken to track short- term changes in the soil. In addition, we assessed the plant community species composition, and measured in situ soil respiration rates.*" Refer "Grassland restoration experiment Kabianga Kenya.pdf" for more info.

## Loading required packages

```{r}
#| eval: true
#| warning: false

library(tidyverse)
library(readxl)
library(GGally)
library(VIM)
library(FactoMineR)
library(lme4)
```

## Loading all datasets

Some unwanted columns have also been removed.

```{r}
#| eval: true
#| warning: false

#Kabianga
kab.nov.22 <- read_excel("data/DSAS_REDEAL_Assign_Data_Aug2.xlsx" , sheet = 'Kab_nov_2022', na = c("NA", NA)) |>
  select(-c(...21,...22))
kab.dec.23 <- read_excel("data/DSAS_REDEAL_Assign_Data_Aug2.xlsx" , sheet = 'Kab_dec_2023', na = c("NA", NA)) 

#Thurgem
thur.jun.22 <- read_excel("data/DSAS_REDEAL_Assign_Data_Aug2.xlsx" , sheet = 'Thur_june_2022', na = c("NA", NA)) |>
  select(-c(...13,...14))

thur.july.23 <- read_excel("data/DSAS_REDEAL_Assign_Data_Aug2.xlsx" , sheet = 'Thur_july_2023', na = c("NA", NA)) |>
  select(-c(...15,...16))

thur.dec.23 <- read_excel("data/DSAS_REDEAL_Assign_Data_Aug2.xlsx" , sheet = 'Thur_dec_2023', na = c("NA", NA))

#Kapsarok
kaps.jun.22 <- read_excel("data/DSAS_REDEAL_Assign_Data_Aug2.xlsx" , sheet = 'Kaps_june_2022', na = c("NA", NA)) |>
  select(-c(...13,...14)) 
```

## Data exploration

#### Missing value check

```{r}
#| eval: true
#| warning: false

aggr(kab.nov.22, prop = F, numbers = T) 
aggr(kab.dec.23, prop = F, numbers = T)
aggr(thur.jun.22, prop = F, numbers = T)
aggr(thur.july.23, prop = F, numbers = T)
aggr(thur.dec.23, prop = F, numbers = T)
aggr(kaps.jun.22, prop = F, numbers = T)

```

For the dataset `kab.nov.23` the plot indicate the last row has all NA entries which is removed. And for the dataset `kaps.jun.22` there are 2 NA entries in the YIELD column, here the average based on the type of treatment for which the value is missing is taken and replaced. Since there are only 2 NA values, code is not written for the entire column.

```{r}
#| eval: true
#| warning: false

kab.nov.22 <- kab.nov.22|>
  na.omit()

mean_yield_1 <- kaps.jun.22 |>
  filter(Code == "S") |>
  pull(YIELD) |>
  mean(na.rm = TRUE)

mean_yield_2 <- kaps.jun.22 |>
  filter(Code == "ST") |>
  pull(YIELD) |>
  mean(na.rm = TRUE)

kaps.jun.22 <- kaps.jun.22 |>
  mutate(YIELD = if_else(Code == "S" & is.na(YIELD), mean_yield_1, YIELD),
         YIELD = if_else(Code == "ST" & is.na(YIELD), mean_yield_2, YIELD)) 
```

#### Checking data types

For the variables `Plot No.`, `YIELD`, `DM`, `SPE`, `Tsoil`, `Tair`, `MOIS`, `RES`, `pH`, `EC`, `MBC`, `MBN`, `Nconc`, `Cconc`, `P`, `BD`, `NH4`, `NO3`, `Block` and for each treatment (later converted to factor type) `M`, `S` and `T` numeric values are expected as entry.

```{r}
#| eval: true
#| warning: false

glimpse(kab.nov.22)
glimpse(kab.dec.23)
glimpse(thur.jun.22)
glimpse(thur.july.23)
glimpse(thur.dec.23)
glimpse(kaps.jun.22)

```

All results where as expected.

#### **Check for multicolinearity**

1.  Kabianga November 2023

```{r}
#| eval: true
#| warning: false
#| fig-width: 12
#| fig-height: 10

kab.nov.22 |>
  select(-`Plot No.`, - Treatment, -M, -S, -T) |> 
  ggpairs()
```

2.  Kabianga December 2023

```{r}
#| eval: true
#| warning: false
#| fig-width: 12
#| fig-height: 10


kab.dec.23 |>
  select(-`Plot No.`, - Treatment, -M, -S, -T) |> 
  ggpairs()
```

3.  Thurgem June 2022

```{r}
#| eval: true
#| warning: false
#| fig-width: 12
#| fig-height: 10

thur.jun.22 |>
  select(-`Plot No.`, - Treatment, -M, -S, -T) |> 
  ggpairs()
```

4.  Thurgem July 2023

```{r}
#| eval: true
#| warning: false
#| fig-width: 12
#| fig-height: 10

thur.july.23 |>
  select(-Plot_No, - Treatment, -M, -S, -T) |> 
  ggpairs()
```

5.  Thurgem Dec 2023

```{r}
#| eval: true
#| warning: false
#| fig-width: 12
#| fig-height: 10

thur.dec.23 |>
  select(- `Plot No.`, - Treatment, -M, -S, -T) |> 
  ggpairs()
```

6.  Kapsorok June 2022

```{r}
#| eval: true
#| warning: false
#| fig-width: 12
#| fig-height: 10

kaps.jun.22 |>
  select(-Plot_No., - Treatment, -M, -S, -T) |> 
  ggpairs()
```

#### **Histogram of YIELD by Treatment faceted by Block**

1.  Kabianga November 2023

```{r}
#| eval: true
#| warning: false

ggplot(kab.nov.23, aes(x = Code, y = YIELD)) +
  geom_bar(color = "black", alpha = 0.7, stat = "identity") +
  labs(title = "Histogram of YIELD by Treaetment faceted by Block (Kabianga Nov 2023)", x = "Treatment", y = "Yield") +
    facet_wrap(~Block) +
  theme_minimal() 

```

2.  Kabianga December 2023

```{r}
#| eval: true
#| warning: false

ggplot(thur.jun.22, aes(x = Code, y = YIELD)) +
  geom_bar(color = "black", alpha = 0.7, stat = "identity") +
  labs(title = "Histogram of YIELD by Treaetment faceted by Block (Kabianga Dec 2023)", x = "Treatment", y = "Yield") +
    facet_wrap(~Block) +
  theme_minimal() 
```

3.  Thurgem June 2022

```{r}
#| eval: true
#| warning: false

ggplot(kab.dec.23, aes(x = Code, y = YIELD)) +
  geom_bar(color = "black", alpha = 0.7, stat = "identity") +
  labs(title = "Histogram of YIELD by Treaetment faceted by Block (Thurgem June 2022)", x = "Treatment", y = "Yield") +
    facet_wrap(~Block) +
  theme_minimal() 
```

4.  Thurgem July 2023

```{r}
#| eval: true
#| warning: false

ggplot(thur.july.23, aes(x = Code, y = YIELD)) +
  geom_bar(color = "black", alpha = 0.7, stat = "identity") +
  labs(title = "Histogram of YIELD by Treaetment faceted by Block (Thurgem July 2023)", x = "Treatment", y = "Yield") +
    facet_wrap(~Block) +
  theme_minimal() 
```

5.  Thurgem Dec 2023

```{r}
#| eval: true
#| warning: false

ggplot(thur.dec.23, aes(x = Code, y = YIELD)) +
  geom_bar(color = "black", alpha = 0.7, stat = "identity") +
  labs(title = "Histogram of YIELD by Treaetment faceted by Block (Thurgem Dec 2023)", x = "Treatment", y = "Yield") +
    facet_wrap(~Block) +
  theme_minimal() 
```

6.  Kapsorok June 2022

```{r}
#| eval: true
#| warning: false

ggplot(kaps.jun.22, aes(x = Code, y = YIELD)) +
  geom_bar(color = "black", alpha = 0.7, stat = "identity") +
  labs(title = "Bar chart of YIELD by Treaetment faceted by Block (Kapsorok June 2023)", x = "Treatment", y = "Yield", caption = "A sample chart based on the dataset Kapsorok June 2023") +
    facet_wrap(~Block) +
  theme_minimal() 
```

## Primary analysis

```{r}
summary(aov(YIELD ~ Block + T * S * M , data = kab.nov.22))
summary(aov(YIELD ~ Block + T * S * M , data = kab.dec.23))
summary(aov(YIELD ~ Block + T * S * M , data = thur.jun.22))
summary(aov(YIELD ~ Block + T * S * M , data = thur.july.23))
summary(aov(YIELD ~ Block + T * S * M , data = thur.dec.23))
summary(aov(YIELD ~ Block + T * S * M , data = kaps.jun.22))

#MBC and MBN
summary(aov(MBC ~ Block + T * S * M , data = kab.nov.22))
summary(aov(MBN ~ Block + T * S * M , data = kab.nov.22))
```

```{At significance level 10%.}
For kab.nov.22  T is significant.
For kab.dec.23 T:S  is significant.


For MBC Block, T, M had significant effect.
For MBN Block and M were. found significant.





For thur.jun.22 no treatment or combinations had a significant effect.
For thur.july.23 T:S is significant.
For thur.dec.23 no treatment or combinations had a significant effect.

For kaps.jun.22 Block, T, S, T:S, T:M and T:S:M had  significance.
```

```{r}
summary(lm(YIELD ~ T * S * M , data = kab.nov.22))
summary(lm(YIELD ~ T * S * M , data = kab.dec.23))
summary(lm(YIELD ~ T * S * M , data = thur.jun.22))
summary(lm(YIELD ~ T * S * M , data = thur.july.23))
summary(lm(YIELD ~ T * S * M , data = thur.dec.23))
summary(lmer(YIELD ~ T * S * M + (1|Block), data = kaps.jun.22))

#MBC and MBN
summary(lmer(MBC ~ T * S * M + (1|Block), data = kab.nov.22))
summary(lmer(MBN ~ T * S * M + (1|Block) , data = kab.nov.22))
```

```{r}
summary(lm(YIELD ~ T * S * M + Block, data = kaps.jun.22))

#MBC and MBN
summary(lm(MBC ~ T * S * M + Block, data = kab.nov.22))
summary(lm(MBN ~ T * S * M + Block, data = kab.nov.22))
```

## Secondary analysis
